---
title: "Quantium Virtual Internship - Retail Strategy & Analytics - Task 1"
author: "Rounak Saha"
date: "2024-11-26"
output:
  pdf_document: default
  html_document: default
---

## Load required libraries and datasets
```{r, echo = TRUE, eval = TRUE, results="hide", fig.keep = "none"}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("tidyverse")
install.packages("readr")
install.packages("readxl")
install.packages("ggplot2")
install.packages("tidyr")
install.packages("ggmosaic")
install.packages("data.table")
```


```{r, echo=TRUE, fig.keep="none", results="hide",warning=FALSE}
library(tidyverse)
library(readxl)
library(readr)
library(ggplot2)
library(tidyr)
library(ggmosaic)
library(data.table)
```

```{r}
# Load the datasets
transactionData <- read_xlsx("QVI_transaction_data.xlsx")
customerData <- fread("QVI_purchase_behaviour.csv")
transactionData <- as.data.table(transactionData)
```
## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look
at each of the datasets provided.

### Examining transaction data
We can use `str()` to look at the format of each column and see a sample of the
data.
We can also run in the console to see a sample of the data or use
`head(transactionData)` to look at the first 10 rows.

```{r}
head(transactionData,10)
```
```{r}
str(transactionData)
```

We can see that the date column is in an numeric format. Let's change this to a
date format.

```{r}
# Convert DATE column to a date format
# A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```

Looks like we are definitely looking at potato chips but how can we check that
these are all chips? We can do some basic text analysis by summarising the
individual words in the product name.
```{r}
prodWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), " ")))
setnames(prodWords, 'names')
```

As we are only interested in words that will tell us if the product is chips or
not, let's remove all words with digits and special characters such as '&' from our
set of product words. We can do this using `grepl()`.
```{r}
prodWords <- prodWords[grepl("\\d", names) ==FALSE, ]
prodWords <- prodWords[grepl("[:alpha:]", names), ]
```

Let's look at the most common words by counting the number of times a word
appears and sorting them by this frequency in order of highest to lowest frequency
```{r}
prodWords[, .N, names][order(-N)]
```

There are salsa and dip products in the dataset but we are only interested in the chips
category, so let's remove these.
```{r}
transactionData <- transactionData[!grepl("salsa|dip", tolower(PROD_NAME)), ]
```

Next, we can use `summary()` to check summary statistics such as mean, min and max
values for each feature to see if there are any obvious outliers in the data and if
there are any nulls in any of the columns (`NA's : number of nulls` will appear in
the output if there are any nulls).
```{r}
summary(transactionData)
```

There are no nulls in the columns but product quantity appears to have an outlier
which we should investigate further. Let's investigate further the case where 200
packets of chips are bought in one transaction.
```{r}
transactionData[PROD_QTY ==200, ]
```

There are two transactions where 200 packets of chips are bought in one transaction
and both of these transactions were by the same customer. 
Let's see if the customer has had other transactions
```{r}
transactionData[LYLTY_CARD_NBR == 226000, ]
```
It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
transactionData <- transactionData[!PROD_QTY== 200, ]
```

```{r}
summary(transactionData)
```

Now, let's look at the number of transaction lines over time to see
if there are any obvious data issues such as missing date.
```{r}
transactionData[, .N, by= DATE ][order(DATE)]
```

There’s only 364 rows, meaning only 364 dates which indicates a missing date. Let’s create a sequence of
dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to
find the missing date.
```{r}
allDays <- as.data.table(seq(min(transactionData$DATE), max(transactionData$DATE), by = 'day'))
setnames(allDays, 'DATE')
allDayTransaction <- merge(allDays, transactionData[, .N, by = DATE], all.x = TRUE)

theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(allDayTransaction, aes(x = DATE, y = N)) + geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over
   time") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + scale_x_date(breaks = "month")
```

We can see that there is an increase in purchases in December and a break in late December. Let’s zoom in
on this.
```{r}
ggplot(allDayTransaction[month(DATE) == 12, ], aes(x = DATE, y = N)) + geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over
   time") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + scale_x_date(breaks = "day")
```

We can see that the increase in sales occurs in the lead-up to Christmas and that
there are zero sales on Christmas day itself. This is due to shops being closed on
Christmas day.

Now that we are satisfied that the data no longer has outliers, we can move on to
creating other features such as brand of chips or pack size from PROD_NAME. We will
start with pack size.

```{r}
#Create Pack Size
transactionData[, packSize := parse_number(PROD_NAME)]

#Let's check if the pack sizes look sensible 
transactionData[, .N, by= packSize][order(-packSize)]
```
The largest size is 380g and the smallest size is 70g - seems sensible!
Let's plot a histogram of PACK_SIZE since we know that it is a categorical
variable and not a continuous variable even though it is numeric.
```{r}
hist(transactionData[, packSize])
```

Pack sizes created look reasonable and now to create brands, we can use the first word in PROD_NAME to
work out the brand name
```{r}
# Extract the first word using tstrsplit
transactionData[, Brand := tstrsplit(PROD_NAME, " ")[[1]]]

transactionData[, .N, by= Brand][order(-N)]
```

Some of the brand names look like they are of the same brands - such as RED and
RRD, which are both Red Rock Deli chips, Smiths and Smith, Doritos and Dorito,Infuzions and Infzns, etc. Let's combine these together.
```{r Clean brand names}
transactionData[, Brand := toupper(Brand)]
#Clean brand names
transactionData[Brand == "RED", Brand := "RRD"]
transactionData[Brand == "SNBTS", Brand := "SUNBITES"]
transactionData[Brand == "INFZNS", Brand := "INFUZIONS"]
transactionData[Brand == "WW", Brand := "WOOLWORTHS"]
transactionData[Brand == "SMITH", Brand := "SMITHS"]
transactionData[Brand == "NCC", Brand := "NATURAL"]
transactionData[Brand == "DORITO", Brand := "DORITOS"]
transactionData[Brand == "GRAIN", Brand := "GRNWVES"]
```

Check again! 
```{r}
transactionData[, .N, by= Brand][order(-N)]
```

### Examining Customer data
Now that we are happy with the transaction dataset, let's have a look at the
customer dataset.We will do some basic summaries of the dataset, including distributions of
any key columns.
```{r}
summary(customerData)
```

```{r}
head(customerData)
```

Let’s have a closer look at the LIFESTAGE and PREMIUM_CUSTOMER columns.
```{r}
customerData[, .N, by= LIFESTAGE]
customerData[, .N, by= PREMIUM_CUSTOMER]
```
As there do not seem to be any issues with the customer data, we can now go ahead and join the transactionand customer data sets together
```{r}
data <- merge(transactionData, customerData, all.x = TRUE)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be
sure that no duplicates were created. This is because we created `data` by setting
`all.x = TRUE` (in other words, a left join) which means take all the rows in
`transactionData` and find rows with matching values in shared columns and then
joining the details in these rows to the `x` or the first mentioned table.

Let’s also check if some customers were not matched on by checking for nulls.
```{r}
sum(is.na(data$LIFESTAGE))
sum(is.na(data$PREMIUM_CUSTOMER))
```

Great, there are no nulls! So all our customers in the transaction data has been
accounted for in the customer dataset.

## Data analysis on customer segments 
Now that the data is ready for analysis, we can define some metrics of interest to
the client:

*Who spends the most on chips (total sales), describing customers by lifestage and
how premium their general purchasing behaviour is

* How many customers are in each segment

* How many chips are bought per customer by segment

* What's the average chip price by customer segment

```{r fig.width = 10, fig.align = "center", warning=FALSE}
#customers by lifestage and how premium their general purchasing behaviour
x <- data[, .(SALES = sum(TOT_SALES)), 
 by = .(PREMIUM_CUSTOMER, LIFESTAGE)][order(-SALES)][, 
 PROP_PERCENT := paste(round(SALES / sum(SALES) * 100, 2), "%")]

# Checking the top 10 purchasing behaviour
head(x, 10)

library(ggmosaic)
# Create plot
p <- ggplot(data = x) +
  geom_mosaic(aes(
    weight = SALES,
    x = product(PREMIUM_CUSTOMER, LIFESTAGE),
    fill = PREMIUM_CUSTOMER
  )) +
  labs(x = "Lifestage", y = "Premium customer flag", title = "Proportion of
 sales") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Plot and label with proportion of sales
p + geom_text(data = ggplot_build(p)$data[[1]], aes(
  x = (xmin + xmax) / 2 ,
  y =
    (ymin + ymax) / 2,
  label = as.character(paste(round(.wt / sum(
    .wt
  ), 3) * 100, '%'))
))
```

Sales are coming mainly from Budget - older families, Mainstream - young
singles/couples, and Mainstream - retirees
Let's see if the higher sales are due to there being more customers who buy chips.
```{r  fig.width = 10, fig.align = "center"}
x1 <- data[, .(CUSTOMER = uniqueN(LYLTY_CARD_NBR)),
 by = .(PREMIUM_CUSTOMER, 
 LIFESTAGE)][order(-CUSTOMER)][, 
 CUST_PERCENT := paste(round(CUSTOMER / sum(CUSTOMER) * 100, 2), "%")]

head(x1,12)

# Create plot
p1 <- ggplot(data = x1) +
  geom_mosaic(aes(
    weight = CUSTOMER,
    x = product(PREMIUM_CUSTOMER, LIFESTAGE),
    fill = PREMIUM_CUSTOMER
  )) +
  labs(x = 'Lifestage', y = 'Customers', title = 'Proportion of Customers') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
# Plot and label with proportion of sales
p1 + geom_text(data = ggplot_build(p1)$data[[1]], aes(
  x = (xmin + xmax) / 2 ,
  y =
    (ymin + ymax) / 2,
  label = as.character(paste(round(.wt / sum(
    .wt
  ), 3) * 100, '%'))
))

```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy
chips. This contributes to there being more sales to these customer segments but
this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.
Let's have a look at this next.
```{r, results = "hide"}
#Average number of units ordered per customer by LIFESTAGE and PREMIUM_CUSTOMER
x2 <- data[, .(AVG_QTY_ORDER = sum(PROD_QTY) / uniqueN(LYLTY_CARD_NBR)), 
  by = .(PREMIUM_CUSTOMER, 
         LIFESTAGE)][order(-AVG_QTY_ORDER)][, 
 ORDER_PERCENT := paste(round(
   AVG_QTY_ORDER / sum(AVG_QTY_ORDER) * 100, 2), "%")]

head(x2,10)

ggplot(data = x2) + geom_col(aes(x = LIFESTAGE, y = AVG_QTY_ORDER, 
                                 fill = PREMIUM_CUSTOMER),
                             position = position_dodge()) +
  labs(x = "Lifestage",
       y = "Average Quantity per Order",
       fill = "Premium Customer",
       title = "Average Quantity by Lifestage and Premium Customer Status") +
  theme_minimal() + ylim(0, 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(panel.border = element_rect(
    color = "black",
    fill = NA,
    size = 1
  ))
```

Older families and young families in general buy more chips per customer.
Let's also investigate the average price per unit chips bought for each customer
segment as this is also a driver of total sales.
```{r}
data[, .(AVG_PRICE_PER_UNIT= sum(TOT_SALES)/sum(PROD_QTY)), 
     by= .(PREMIUM_CUSTOMER,LIFESTAGE)][order(-AVG_PRICE_PER_UNIT)]

p3 <- ggplot(data = data[, .(AVG_PRICE_PER_UNIT = sum(TOT_SALES) / sum(PROD_QTY)), 
  by = .(PREMIUM_CUSTOMER, LIFESTAGE)][order(-AVG_PRICE_PER_UNIT)]) +
  geom_col(aes(x = LIFESTAGE, y = AVG_PRICE_PER_UNIT, fill = PREMIUM_CUSTOMER),
           position = position_dodge()) +
  labs(x = "Lifestage",
       y = "Avg price per unit",
       fill = "Premium Customer",
       title = "Price per unit") +
  theme_minimal() + ylim(0, 4.5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

p3 + theme(
  panel.border = element_rect(color = "black", fill = NA, size = 1)
)
```

Mainstream midage and young singles and couples are more willing to pay more per
packet of chips compared to their budget and premium counterparts. This may be due
to premium shoppers being more likely to buy healthy snacks and when they buy
chips, this is mainly for entertainment purposes rather than their own consumption.
This is also supported by there being fewer premium midage and young singles and
couples buying chips compared to their mainstream counterparts.

As the difference in average price per unit isn't large, we can check if this difference is statistically different.
Perform an independent t-test between mainstream vs premium and budget midage
and young singles and couples.
```{r}
# Two sample t-test
data[, PRICE_PER_UNIT := TOT_SALES / PROD_QTY]
#First-sample
sample1 <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                  PREMIUM_CUSTOMER == "Mainstream", PRICE_PER_UNIT]
#Second-sample
sample2 <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                  PREMIUM_CUSTOMER != "Mainstream", PRICE_PER_UNIT]

t.test(sample1, sample2, alternative = "greater")
```

The t-test results in a p-value < 2.2e-16, i.e. the unit price for mainstream,
young and mid-age singles and couples are significantly higher than
that of budget or premium, young and mid-age singles and couples.

## Deep dive into specific customer segments for insights
We might want to target customer segments that contribute the most to sales to
retain them or further increase sales. Let's look at Mainstream - young
singles/couples. For instance, let's find out if they tend to buy a particular
brand of chips.

```{r}
# Deep dive into Mainstream, young singles/couples
segment1 <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                   PREMIUM_CUSTOMER == "Mainstream", ]

other <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                PREMIUM_CUSTOMER != "Mainstream", ]
# Brand affinity compared to the rest of the population
segment1_qty <- segment1[, sum(PROD_QTY)]
other_qty <- other[, sum(PROD_QTY)]

seg1_qty_prop_byBrand <- segment1[, .(targetSegment = sum(PROD_QTY) / segment1_qty), by = Brand]
other_qty_prop_byBrand <- other[, .(other = sum(PROD_QTY) / other_qty), by = Brand]

affinity <- merge(seg1_qty_prop_byBrand, other_qty_prop_byBrand)[, affinityToBrand :=
                                                                   targetSegment /                                                                  other][order(-affinityToBrand)]

affinity
```
We can see that :

* Mainstream young singles/couples are 28% more likely to purchase KETTLE chips compared to the
rest of the population

* Mainstream young singles/couples are 58% less likely to purchase Burger Rings compared to the rest of the population

Let's also find out if our target segment tends to buy larger packs of chips.
```{r}
# Deep dive into Mainstream, young singles/couples
segment1 <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                   PREMIUM_CUSTOMER == "Mainstream", ]

other <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                PREMIUM_CUSTOMER != "Mainstream", ]
# Brand affinity compared to the rest of the population
segment1_packSize <- segment1[, sum(PROD_QTY)]
other_packSize <- other[, sum(PROD_QTY)]

seg1_qty_prop_byPackSize <- segment1[, .(targetSegment = sum(PROD_QTY) /
                                           segment1_packSize), by = packSize]
other_qty_prop_byPackSize <- other[, .(other = sum(PROD_QTY) / other_packSize), 
                                   by = packSize]

packSizeProportion <- merge(seg1_qty_prop_byPackSize, 
                            other_qty_prop_byPackSize)[, packSizeProp := 
                                                         targetSegment /other][order(-packSizeProp)]

packSizeProportion
```
It looks like Mainstream young singles/couples are 36% more likely to purchase a 330g pack of chips compared to the rest of the population but let’s dive into what brands sell this pack size.
```{r}
data[packSize==330, unique(PROD_NAME)]
```

Doritos and Smiths are the only brand offering 330g packs and so this may instead be reflecting a higher likelihood of purchasing these.

## Conclusion

* Sales have mainly been due to Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees shoppers. We found that the high spend in chips for mainstream young singles/couples and retirees is due to there being more of them than other buyers. 

* Mainstream, midage and young singles and couples are also more likely to pay more per packet of chips. This is indicative of impulse buying behaviour.

* We’ve also found that Mainstream young singles and couples are 28% more likely to purchase KETTLE chips compared to the rest of the population. The Category Manager may want to increase the category’s performance by off-locating some KETTLE and smaller packs of chips in discretionary space near segments
where young singles and couples frequent more often to increase visibilty and impulse behaviour.
















